Exploring the conformational space of a large molecule in solution using an external electric field: A Nonequilibrium Molecular Dynamics Simulation

### 2013-08-20 writedown/npt-neq/alanine.tex
Luigi's revise

\title{Exploring Conformational Dynamics and Related Time Scales via Nonequilibrium Molecular Dynamics Simulation}
% \title{The non-equilibrium molecular dynamics simulation of the alanine dipeptide under electric field}
\author{Han Wang}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin, Germany}
\author{Giovanni Ciccotti}
\affiliation{Department of Physics, University of Roma "La Sapienza", Italy}
\author{Luigi Delle Site}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

The possible effect of the electric field (EF) on the conformation of
proteins gains increasing attention recently, because it is closely
related to the health problem of exposing human tissue in the
electromagnetic radiation. Despite the importance of this problem, a
complete understanding of how the EF influence the  protein
is still lacking, therefore, a considerable number of
experimental~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects} studies have been devoted to this topic,
and the references listed above is far from complete.

The molecular dynamics (MD) simulation has been proved to be useful in
understanding the behavior of a protein (or a short peptide segment)
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB).
And then the temperature controlled simulation are performed for 
a few to tens of nanoseconds, during which the interested
conformational changes can be observed. Along the trajectories, the
secondary structure, root mean square displacement (RMSD), the dipole
moment and some shape parameters are calculated and investigated by
either plotted as a function of time, or doing time average over the
whole simulation. This approach release useful information on how the
secondary structure is changed by external  EF.  However,
by plotting the quantities against time, one may observe strong
fluctuation in plots, which hintered the quantitative and conclusive
analysis on what is the trend of the conformational change, when
it happens and what is the strength.
Although the time average reduces the statistical uncertainty,
it is indeed not well defined for a nonequilibrium process
such as relaxing process to the new conformations under a static
EF, or the dynamics under an oscillating EF.

The dynamical nonequilibrium molecular dynamics simulation (D-NEMD)
was firstly developed by G. Ciccotti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought}, and
recently successfully applied to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. D-NEMD provides a way of evaluating
nonequilibrium observables by doing an ensemble of nonequilibrium MD
simulations.  It can be used for analyzing the
conformational changing under a generally time-dependent EF.
In the present paper, we study a small model peptide
segment -- alanine dipeptide. We take the approach of starting from a
single peptide piece rather than a more realistic chain, firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding all possible interplay between
different peptide segments along a chain.
Secondly, the current computational resource available is not enough
to study peptide chains by D-NEMD, since one needs to simulate an
ensemble of MD trajectories (typically a few thousands), so it
requires much higher computational cost than producing a single
equilibrium MD trajectory.

In this paper, we firstly describe the idea of the D-NEMD, then
introduce the
observables for analyzing the nonequilibrium properties
(i.e., the dipole moment, time-dependent properties of being a certain
conformation and the net
probability flux). The numerical results of alanine
dipeptide under static and sine-shaped oscillating EF are
presented. By calculating the nonequilibrium properties from the D-NEMD
simulations, we are able to answer how the peptide
changes the conformation in a EF, when do these changes happens,
and how strong are they.
As far as we know, our simulation is the first D-NEMD simulation
of alanine dipeptide under EF, and provides new
information helping people better understand the nonequilibrium
effect of EF on the protein conformation.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
Here we fastly review the approach of performing dynamical nonequilibrium
molecular dynamics (D-NEMD) that was initiated by Giovanni Ciccoti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
We denote the macroscopic nonequilibrium observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
where $\hat O (\vect x)$ is the microscopic observable, 
which is measured at the phase space position $\vect x$,
comparing with the macroscopic observable that is not phase space independent.
In this work, we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known, and is identical to the equilibrium distribution
without any EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes the inner produce in the
phase space.  We assume the dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x)$, where $\mathcal H$ is the Hamiltonian, and $J$ is
the symplectic matrix, then the Liouville equation of the probability
distribution is
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = e^{-iL(t)} \rho(\vect x, 0)$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by $\hat O(\vect x(t)) = e^{iL(t)} \hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), e^{-iL(t)} \rho(\vect x, 0)\rangle
  = \langle e^{iL(t)}\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the equilibrium
distribution, an
explicit translation of Eq.~\eqref{eqn:tmp4} is that the
nonequilibrium observable is equal to the ensemble average of
microscopic observable measured along trajectories, the
initial configurations of which are subject to the \emph{equilibrium}
distribution. In practice, we firstly run an equilibrium MD simulation
to generate a sampling of configurations that are subject to the
equilibrium distribution. And then by using these configurations as
initial configuration, the Hamiltonian dynamics are integrated until
time $t$. These trajectories are also called \emph{branching
  trajectories} in this paper. Finally, the macroscopic observable is
estimated by averaging the microscopic observable measured at the end
points of the trajectories.

Being more generalized, if the microscopic observable depends not only on the end
configuration at time $t$, but also on the historical
configurations along the trajectory, then expression~\eqref{eqn:tmp4} takes the generalized form:
\begin{align}
  O(t) = \int d\vect x\,\rho(\vect x, 0) \int_{\mathcal C\{\vect x, 0; t\}} \hat O[\vect x_s] \,d\mathcal P[\vect x_s] 
\end{align}
where the microscopic observable $ \hat O[\vect x_s] $ is now a
functional of the trajectory.  $\mathcal C\{\vect x, 0; t\}$ is the
set of all continuous trajectories starting at point $\vect x$ at time
0, and ending at time $t$. $\mathcal P[\vect x_s] $ is the measure of
the trajectory space $\mathcal C$.  In our case, it is the
$\delta$-function peaked around the trajectories governed by the
Hamiltonian dynamics. To be general, it can also be the measure of all
trajectories generated by, for example, a Langevin dynamics, then the
branching trajectories should be generated by the Langevin
dynamics rather than the Hamiltonian dynamics.

\subsection{Nonequilibrium temperature control}\label{sec:tmp2b}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig/thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control.
    The white part is the Hamiltonian dynamics region
    and the gray part is thermostating region.
    $R_{ex}$ denotes the size of the Hamiltonian dynamics region, which
    is a sphere located at the center of the box in the Figure.
    The whole system
    is subject to the periodic boundary condition. }
  \label{fig:tmp2}
\end{figure}

% The aforementioned D-NEMD simulation algorithm
% does not describe how the boundary of the system should be set up.
% The very naive idea of doing the simulation
% is to simulate a infinitely large isolated system, or any large enough
% isolated subsystem of it.
% In practice, 
% Therefore, we are interested in a

We are mainly interested in the
nonthermal~\cite{delaHoz2005microwaves} effect of the EF
to the configuration of a solvated alanine dipeptide,
so the temperature control in a nonequilibrium system
becomes an important issue, becuase it was argued that  an
accurate internal reaction temperature control is essential for being able to
perform reproducible experiments~\cite{damm2012can}.
A naive (but close to realistic) implementation of the temperature control is to 
embeded the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine, so that
any extra heat generated by the EF can be effectively
absorbed by the solvent environment.
However, in practice, it is sometime too expensive to explicitly
simulate a very large solvent environment
due to the usually limited computational resources.
Therefore, we are interested in how to cheaply replace the infinitely large
solvent environment by a small affordable system, while
at the same time accurately reproduce the nonequilibrium
physical properties of interest.
% In practice, it is impossible to simulate an infinitely large
% environment.
If only the equilibrium properties are interested, the solution is to couple
the system to a thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by only simulating a finite size system with periodic boundary
condition.
However, directly coupling a thermostat to the nonequilibrium system
will disturb the original Hamiltonian dynamics. This is
in general problematic, 
because the dynamics  under thermostating is artificial designed,
so that the equilibrium distribution can be correctly produce, but not
the nonequilibrium properties of system.
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% However, for a nonequilibrium system, the \emph{dynamics} properties
% will be distrubed by the thermostat, because the
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% For a nonequilibrium simulation, simply couple the system
% to a thermostat may not be a good idea, because the \emph{dynamics}
% generated by the thermostat is artificial, and deviation from the
% Hamiltonian dynamics under investigation will introduce artificial
% effects in the nonequilibrium observations.
To solve this problem, we
observe that only preserving the Hamiltonian dynamics of the alanine 
itself and the water molecules in the nearby solvation shell is
crucial to the interested nonequilibrium observable
(in this work, it is the conformational change of alanine),
while the detailed dynamics of
the water molecules far away is of less importance. 
Therefore, instead of simulating an infinitely
large system, we only simulate a finite size system with periodic
boundary condition. Accordingly, the simulating region is divided into two
subregions: Near the alanine dipeptide, the dynamics is kept to be
Hamiltonian, and this subregion is called \emph{the Hamiltonian dynamics
region}. While far from the alanine, the dynamics of water is coupled
to a Langevin thermostat, and this region is called \emph{the thermostating
region} (see Fig.~\ref{fig:tmp2}).  For simplicity, the Hamiltonian
dynamics region is assumed to be spherical, centered at the
alpha-carbon of the alanine, and with radius of $R_{ex}$.  Further, we
freeze the movement of the alpha-carbon, so that the alanine is always
located at the center of the simulation region.
Therefore, the dynamics is preseved in the Hamiltonian dynamics region (where the properties of interest are observed)
and the artificial effect of thermostating is negligible due to the finite corrlation
range of liquid water.
At the same time, the thermostating region works as a infinitely large
environment that effectively absorbs the extra heat, so that
the nonthermal effect can be studied.
The validity of the above
statement will be later check latter by numerical examples showing that the
nonequilibrium observables do not depends on the size of the system
and the size of the Hamiltonian dynamics subregion, if they are
reasonably large and the finite size effect does not play a dominating role.

@@

\section{Example I: Alanine dipeptide
  under a constant EF}

\subsection{System settings and simulation protocol}

In this section,
we want to study the nonequilibrium properties of an alanine dipeptide
under a constant EF.  At time $t=0$~ps, the system
has been fully equilibriated without any EF. From $t=0$ to
warm-up time $t=t_{\textrm{warm}}$, the eclectic field is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of this field is arbitrarily chosen: along the
$x$ direction. Since the dipole of the equilibrium configurations of
alanine can be along any direction, applying the electric along $x$-axis is
not biased.
Being more general, in this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, the constant EF is
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ as $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
as $t \geq t_{warm}$.
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and solved in 644 TIP3P
water molecules.
The EF exserts force $\vect F = q \vect E(t)$ on
the partial charge $q$ in the system.
The size of the  Hamiltonian dynamics region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
Firstly, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$) coupled to the system for constant
temperature.  Along the trajectory, configurations were taken every
50~\textsf{ps}, so we use 2000 initial configurations for each nonequilibrium
MD simulation, if not stated otherwise.
% They server as starting point of the branching
% trajectories.
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by Gromacs.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostating region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at the ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an obviously effect on the
dynamics of the system.

@@

\begin{figure}
  \centering
  \raisebox{0.2\height}{\includegraphics[width=0.22\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/c-2.eps}}
  \includegraphics[width=0.37\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/fig-begin-1.eps}
  \includegraphics[width=0.37\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/fig-end-1.eps}
  \caption{The population of metastable states of the alanine dipeptide.
    (a): the
    equilibrium state at $t=0$~\textsf{ps}. (b): The new equilibrium
    state at $t=1000$~\textsf{ps}, i.e., fully relaxed under the
    constant EF 1~V/nm. The darker color
    indicates larger population on the Ramachandran histogram.}
  \label{fig:tmp4}
\end{figure}

>>>>

\title{Exploring the conformational space of a large molecule in solution using an external electric field: A Nonequilibrium Molecular Dynamics Simulation}
% \title{The non-equilibrium molecular dynamics simulation of the alanine dipeptide under electric field}
\author{Han Wang}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin, Germany}
\author{Giovanni Ciccotti}
\affiliation{Department of Physics, University of Roma "La Sapienza", Italy}
\author{Luigi Delle Site}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

The possible effect of the electric field (EF) on the conformation of
proteins is a subject of increasing attention since it is linked to the problem of exposing human tissues to
electromagnetic radiation. Despite its major relevance, a
satisfactory understanding of how the EF influences the behaviour of proteins
has not been reached so far although a considerable number of
experimental~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects} studies have been devoted to this topic;
where the references reported above represent only a minor sample of examples.

In this perspective, molecular dynamics (MD) has been proved to be useful in
understanding the behavior of a protein 
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB),
and then temperature controlled simulations are performed for a maximum extension of the order of 10-60 of nanoseconds.
During this simulation time 
relevant conformational changes characterizing the solvated molecule can be observed. 
Along the trajectories, various observable, such as the
secondary structure, root mean square displacement (RMSD), the dipole
moment, shape parameters are calculated as a function of time or taking its time average. This approach releases useful information on how the
secondary structure of the molecule is changed by external  EF.  However,
by plotting some quantities against time, one may observe strong
fluctuations; the entity of such fluctuations are such that a quantitative 
analysis becomes difficult and does not allow for  a clear understanding of the rationale behind the conformational changes.
Although the time average reduces the statistical uncertainty,
however in this specific case, it is indeed not well defined;
in fact we are in the presence of a nonequilibrium process
(e.g. relaxing process to the new conformations under a static
EF, or the dynamics from one conformation to the other under an oscillating EF).

In this perspective, if one wants to use MD, the proper way to do it would be that of treating a situation of non equilibrium with algorithms that can explicitly describe it. This is the approach that will be followed in this work.

The method we employ is the dynamical nonequilibrium molecular dynamics simulation (D-NEMD).
The method was firstly developed by one of us (GC) and coworkers~\cite{ciccotti1975direct, ciccotti1979thought}.
It has been recently applied, with success, to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. The essential feature of D-NEMD is the following: It provides a way of evaluating
nonequilibrium observables by using a properly determined ensemble of nonequilibrium MD
simulations.  In this context, the method can be used for analyzing the
conformational changes of a solvated molecule, subject to a time-dependent EF.
In the present paper, we study a small peptide, alanine dipeptide, as a model for showing how the method can be used. 
The choice of a small peptide, rather than a larger one, is done firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding a possible interplay between
different peptide segments along a chain.
Secondly, the computational cost required is rather massive and thus the testing of the various methodolgy developments proposed here would not be easy to accomplish for large molecules. Of course, given a positive coming out for such a model system, the aim is that of approaching systems of growing complexity.
The paper is organized as follows:
We firstly describe the idea of D-NEMD, then
introduce the physical
observables for analyzing the nonequilibrium properties. The numerical results of alanine
dipeptide under static and sine-shaped oscillating EF are
finally presented. By calculating the nonequilibrium properties from D-NEMD
simulations, we are able to address the question of how the peptide
changes the conformation in a EF (conformational space), when do these changes happens (time scale),
and how likely they are (probability). 
To our knowledge, this study is one of the few (if not the first)  where non equilibrium is explicitly considered for a large molecule in solution bubject to time-dependent EF.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
In this section we briefly review the approach for performing dynamical nonequilibrium
molecular dynamics (D-NEMD) \cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
In the following we denote the macroscopic observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
we refer to $\hat O (\vect x)$ as the microscopic observable, 
which is measured at each point in the phase space $\vect x$.
Here we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known and is identical to the equilibrium distribution of the system
without EF.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes, as usual, the inner product in the
phase space.  We assume tthat he dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x)$, where $\mathcal H$ is the Hamiltonian, and $J$ is
the symplectic matrix. The Liouville equation for the probability
distribution writes:
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = e^{-iL(t)} \rho(\vect x, 0)$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by $\hat O(\vect x(t)) = e^{iL(t)} \hat O(\vect x(0))$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), e^{-iL(t)} \rho(\vect x, 0)\rangle
  = \langle e^{iL(t)}\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the system starts from the distribution of equilibrium (without EF)
 Eq.~\eqref{eqn:tmp4} expresses the fact that the 
observable $O(t)$ calculated in a situation of nonequilibrium is equal to the ensemble average of
the microscopic observable measured along trajectories, the
initial configurations of which are subject to the \emph{equilibrium}
distribution. In practical terms we proceed by first running an equilibrium MD simulation
to generate a sample of configurations. Next we employ these configurations as
initial configurations and, for each, the Hamiltonian dynamics is integrated until
time $t$. These trajectories are also called \emph{branching
  trajectories} in this paper (see Fig.\ref{branching} for a pictorial representation). 
Finally, the value of the macroscopic observable at time $t$ is
calculated by averaging the microscopic observable measured at each time $t$ along each branching trajectory.

%Being more generalized, if the microscopic observable depends not only on the end
%configuration at time $t$, but also on the historical
%configurations along the trajectory, then expression~\eqref{eqn:tmp4} takes the generalized form:
%\begin{align}
%  O(t) = \int d\vect x\,\rho(\vect x, 0) \int_{\mathcal C\{\vect x, 0; t\}} \hat O[\vect x_s] \,d\mathcal P[\vect x_s] 
%\end{align}
%where the microscopic observable $ \hat O[\vect x_s] $ is now a
%functional of the trajectory.  $\mathcal C\{\vect x, 0; t\}$ is the
%set of all continuous trajectories starting at point $\vect x$ at time
%0, and ending at time $t$. $\mathcal P[\vect x_s] $ is the measure of
%the trajectory space $\mathcal C$.  In our case, it is the
%$\delta$-function peaked around the trajectories governed by the
%Hamiltonian dynamics. To be general, it can also be the measure of all
%trajectories generated by, for example, a Langevin dynamics, then the
%branching trajectories should be generated by the Langevin
%dynamics rather than the Hamiltonian dynamics.

\subsection{Control of temperature in Nonequilibrium MD}\label{sec:tmp2b}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig/thermostat.eps}
  \caption{Schematic plot of the nonequilibrium temperature control scheme.
    The white area is the region characterized by Hamiltonian dynamics without the direct action of an external thermostat.
    The gray area is the region where the external thermostat is used in order to keep the target temperature.
    $R_{ex}$ denotes the size of the Hamiltonian dynamics region, which
    is a sphere at the center of the box.
    The whole system is subject to the periodic boundary conditions. }
  \label{fig:tmp2}
\end{figure}

% The aforementioned D-NEMD simulation algorithm
% does not describe how the boundary of the system should be set up.
% The very naive idea of doing the simulation
% is to simulate a infinitely large isolated system, or any large enough
% isolated subsystem of it.
% In practice, 
% Therefore, we are interested in a 
For this study We are interested in 
nonthermal~\cite{delaHoz2005microwaves} effects of the EF
on the configuration of the solvated alanine dipeptide.
For this reason a proper control of the temperature in a nonequilibrium system
becomes an important issue; in fact it was argued that  an
accurate control of the internal reaction temperature is essential for being able to
perform reproducible experiments~\cite{damm2012can}.
A straigthforward (and realistic) implementation of the temperature control consists in 
embedding the alanine into an infinitely large solvent environment, and 
the EF is only applied to the neighborhood of the alanine. In this case
any extra heat generated by the EF can be effectively
dispersed in the large solvent environment.
In this way one can identify the effect of the EF on the conformational space of the solvated molecule,
without the artifact that the surrounding solvant cannot release the additional energy of the EF.
However, in practice, in most of the cases the computational cost of having large solvant environment is rather massive if not prohibitive.
Therefore, it is of interest to substitute the infinitely large
solvent environment with a smaller affordable system and at the same time accurately reproduce the nonequilibrium
physical properties of interest.

As it is well known, in MD, for equilibrium situation, the solution is to couple
the system to a an external thermostat, under which the desired ensemble
(canonical in this case) can be
sampled by simulating a finite size system with periodic boundary
conditions.
However, coupling the system to a thermostat for non equilibrium situations is a delicate issue.
In fact, since a thermostat is explicitly designed to bring the system at equilibrium, any perturbation produced by the EF, would be mostly adsorbed by the thermostat. This means that the physics of the original Hamiltonian dynamics with the EF would be lost and the response to the perturbation becomes unphysical. 
%will disturb the original Hamiltonian dynamics. This is
%in general problematic, 
%because the dynamics  under thermostating is artificial designed,
%so that the equilibrium distribution can be correctly produce, but not
%the nonequilibrium properties of system.
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% However, for a nonequilibrium system, the \emph{dynamics} properties
% will be distrubed by the thermostat, because the
% dynamics of the system is no longer Hamiltonian, and is
% artificial designed to reach the equilibrium probability distribution.
% For a nonequilibrium simulation, simply couple the system
% to a thermostat may not be a good idea, because the \emph{dynamics}
% generated by the thermostat is artificial, and deviation from the
% Hamiltonian dynamics under investigation will introduce artificial
% a effects in the nonequilibrium observations.
In this section we propose a procedure to solve the problem: we
observe the Hamiltonian dynamics (without thermostat) of the alanine 
itself and the water molecules in the closer solvation shells is
crucial in order to determine the response of the observable of interest
under the perturbation of the EF (in this work, it is the conformational change of alanine).
Instead the detailed dynamics of
the water molecules far away is not of primary importance, that is the region beyond the first solvation shells plays the role of a thermodynamic bath only. 
Therefore, instead of simulating an infinitely
large system, we only simulate a finite size system with periodic
boundary condition with the following characteristics. The simulating region is divided in two
subregions: a spherically defined region around the alanine dipeptide, centered at the
alpha-carbon of radius $R_{ex}$. Here the dynamics is
Hamiltonian and molecules are not subject to a thermosat; we refer to such subregion as \emph{the Hamiltonian dynamics
region}. Instead beyond $R_{ex}$, the dynamics of water is coupled
to a Langevin thermostat, and this region is called \emph{the thermostating
region} (see Fig.~\ref{fig:tmp2}). For technical convenience we
fix the position of the alpha-carbon in space, so that the alanine is always
located at the center of the simulation region.
Therefore, a realistic non equilibrium  dynamics is preseved in the Hamiltonian dynamics region (where the properties of interest are observed)
and possible direct artificial effects of the thermostating process in the region outside are negligible because of the finite correlation
range of liquid water.
At the same time, the thermostating region works as a infinitely large
environment that effectively absorbs the extra heat produced by the EF.
The validity of the above
statement will be later on carefully checked by a series of numerical tests where it is shown that
the response of observables in nonequilibrium (i.e. under the effect of the EF) does not depend on the size of the system
and on the size of the Hamiltonian dynamics subregion, provided that they are
reasonably large.

@@

\section{Example I: Alanine dipeptide
  under a uniform  constant EF}

\subsection{System settings and simulation protocol}
At time $t=0$~ps, the system
has been fully equilibriated without any EF. From $t=0$ to
warm-up time $t=t_{\textrm{warm}}$, the eclectic field is switched on linearly, while
after $t=t_{\textrm{warm}}$, the field is kept constant in time at
$E_{\infty}$.  In this work we consider $t_{\textrm{warm}} = 10$~ps,
and $E_{\infty} = 1$~V/nm.
The direction of the field is arbitrarily chosen along the
$x$ direction, in fact the dipole of the equilibrium configurations of
alanine can be along any direction.
In this paper we denote the
EF as a function of time $\vect E(t)$.
Therefore, the constant EF is
$\vect E(t) = (E_\infty\cdot t/t_{warm},0,0)$ as $0\leq t < t_{warm}$, and 
$\vect E(t) = (E_\infty,0,0)$
as $t \geq t_{warm}$.
The system is set up in a $2.7\times 2.7\times
2.7\, \textsf{nm}^3$ periodic simulation region, with one alanine dipeptide
described by the CHARMM27 force field, and solved in 644 TIP3P
water molecules.
%The EF exserts force $\vect F = q \vect E(t)$ on
%the partial charge $q$ in the system.
The size of the  Hamiltonian dynamics region is $R_{ex} = 1.0$~nm.
All simulations are performed by a home-modified GROMACS 4.5~\cite{pronk2013gromacs}.
First, an equilibrium NVT simulation of
100~\textsf{ns} was performed with a Langevin thermostat (time-scale
$\tau_T = 0.5~\textsf{ps}$).  Along the trajectory, configurations were taken every
50~\textsf{ps} and we used 2000 initial configurations for each nonequilibrium
MD simulation (if not stated otherwise for specific cases).
% They server as starting point of the branching
% trajectories.
The branching trajectories were integrated by the
Leap-frog scheme with the aforementioned nonequilibrium
temperature control technique.  The
time step was $\Delta t = 0.002~\textsf{ps}$. The short-range
interaction was smoothed from $0.8$ to $1.0~\textsf{nm}$ by the
``switch'' method provided by Gromacs.  An energy conserving PME
method was applied to calculate the electrostatic interaction in this
periodic system. In the thermostating region, the original dynamics was
coupled to a Langevin thermostat with $\tau_T = 0.1~\textsf{ps}$.
The whole system is also coupled to a Parrinello-Rahman barostat with $\tau_P = 2.0~\textsf{ps}$ to keep
the pressure at the ambient condition (1~Bar). Since the
change of the system size is small and slow, the pressure control
does not have an sizeable effect on the
dynamics of the system.

@@

\begin{figure}
  \centering
  \raisebox{0.2\height}{\includegraphics[width=0.22\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/c-2.eps}}
  \includegraphics[width=0.37\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/fig-begin-1.eps}
  \includegraphics[width=0.37\textwidth]{fig.ala/ext.mode1.010.Ex.01.00.t1000ps.recheck/fig-end-1.eps}
  \caption{The population of metastable states of the alanine dipeptide.
    (a): the
    equilibrium state at $t=0$~\textsf{ps}. (b): The new equilibrium
    state at $t=1000$~\textsf{ps}, i.e., fully relaxed under the
    action of a constant EF of 1~V/nm. The darker color
    indicates larger population on the Ramachandran histogram.}
  \label{fig:tmp4}
\end{figure}

For each branching trajectory,  the molecular configuration of alanine is projected on
the Ramachandran histogram, as
shown in Fig.~\ref{fig:tmp4}. The equilibrium distribution of the
conformations is given in the plot (a), and the fully relaxed
conformation under the action of $E_{\infty} = 1$~V/nm is given  plot (b).  It is
evident from the plot that the conformations are grouped in several clusters, and the
position of the clusters does not change when the external electric
field is applied.
{\bf LUIGI COMMENT: HAN below is really not clear}\\
This means that only few of the total of conformations are likely to be visited,
in the nonequilibrium process.
These conformations are called ``metastable states''.
The strict definition of the
metastability is not exactly the same as those conformations that are likely to be observed.
Since in this paper we do not focus on the discussion of the definition for the metastability of the system,
we do not distinguish the definations, and assume they have the same meaning.
{\bf END COMMENT}\\
For sake of clarity We divide the
Ramachandran histogram into 5 subregions, see Fig.~\ref{fig:tmp4},
then the analysis of conformations on the Ramachandran histogram is reduced to the
analysis of the 5 states, i.e. $\{A_1, A_2, B_1, B_2, C\}$.
These states
correspond to different secondary structure of a peptide chain:
$A_1$ and $A_2$ are
corresponding to the alpha helix. $B_1$ and $B_2$ are
corresponding to the beta sheet. $C$ is corresponding to the
left-handed alpha helix.
